---
title: "GAMM analyses"
output:
  html_document: default
  word_document: default
  pdf_document: default
date: "2026-04-20"
---



```{r,include = FALSE}
library(dplyr)

df <- read.csv("/Users/yangji/Downloads/osfstorage-archive/final_merged_manybabies.csv", header = TRUE, sep = ",")

# Only variables specified for habituation analyses in the variable dictionary are included.

df=df[df$habituation==TRUE,]


```

## Data preparations

In GAMM analyses, several key variables are required. Primarily, the time-series analysis utilizes Y and X derived from the `looking_time` and `trial_number` variables. A new measure, `looking_time2`, was created to incorporate looking time during the freeze-frame period.

The primary focus of this analysis is the experimental conditions. For this MBE, these are nested within `trial_type` and have been divided into three new variables: infant-directedness (`tt_dir`), consistency (`tt_con`), and social context (`tt_soc`). Any null values have been recoded as dummy values.

Interaction variables include `method` and `participant_sex`. To account for developmental trends, we can utilize either the continuous variable `age_days` or the discrete variable `age_group_new`. While `stimulus_id` could be recoded to further investigate specific stimulus types, this is likely unnecessary for the current analysis.

The random effect is accounted for by the `unique_subject_identifier` variable.

*Exclusions:*

- `stimulus_type` was excluded as it contains only one unique value.

- `project_id` and `lab_id` were excluded because they represent higher-level groupings than `trial_type` and `unique_subject_identifier`, respectively.

- Demographic variables are not used, since they are controlled altogether by `unique_subject_identifier`, unless we need to look into these details.


<br>

### Overview

```{r}
names(df)
```

<br>

#### Y: `looking_time` (numeric)

A supplemental variable, `looking_time2`, was created by incorporating looking time from the MB4 freeze-frame period into the conventional looking time measure.

```{r}

df$looking_time2=0
df$looking_time2[!is.na(df$lookingtime_freezeframe)]=df$lookingtime_freezeframe[!is.na(df$lookingtime_freezeframe)]
df$looking_time2=df$looking_time+df$looking_time2


```

<br>

#### X: `trial_number` (numeric)

```{r,echo=FALSE}

df$trial_number <- as.numeric(df$trial_number)
df <- df[df$trial_number <= 12,]
paste("MB1: ",paste(sort(unique(df$trial_number[df$project_id=="MB1"])),collapse=", "),sep="")
paste("MB3: ",paste(sort(unique(df$trial_number[df$project_id=="MB3"])),collapse=", "),sep="")
paste("MB4: ",paste(sort(unique(df$trial_number[df$project_id=="MB4"])),collapse=", "),sep="")
```

<br>

#### Exp: `trial_type` (ordered, with dummy)

The `trial_type` column contains nested conditions for the MB1, MB3, and MB5 sub-tasks, which required recategorization for analysis. For instance, to examine the infant-directedness aspect of MB1, a new variable, `tt_dir`, was created to distinguish between 'IDS' (Infant-Directed Speech) and 'ADS' (Adult-Directed Speech). Observations from MB3 and MB5, which did not involve these conditions, were assigned `NA` values.

*Exclusions*: The higher-level variable `project_id` was no longer required.

```{r}
unique(df$trial_type)
```

```{r,include = FALSE}
df$tt_con <- ifelse(df$trial_type %in% c("consistent","inconsistent"),
                    df$trial_type, NA)
df <- df %>%
  mutate(tt_con = case_when(
    is.na(tt_con) ~ 0,
    tt_con == "consistent" ~ 1,
    tt_con == "inconsistent" ~ 2
  ))

df$tt_soc <- ifelse(df$trial_type %in% c("nonsocial","social"), 
                    df$trial_type, NA)
df <- df %>%
  mutate(tt_soc = case_when(
    is.na(tt_soc) ~ 0,
    tt_soc == "social" ~ 1,
    tt_soc == "nonsocial" ~ 2
  ))

df$tt_dir <- ifelse(df$trial_type %in% c("IDS","ADS"), 
                    df$trial_type, NA)
df <- df %>%
  mutate(tt_dir = case_when(
    is.na(tt_dir) ~ 0,
    tt_dir == "IDS" ~ 1,
    tt_dir == "ADS" ~ 2
  ))

df$tt_dir=as.ordered(df$tt_dir)
df$tt_con=as.ordered(df$tt_con)
df$tt_soc=as.ordered(df$tt_soc)
```

To facilitate contrast analyses with GAMM, these categories were converted into ordered variables where NA is coded as 0, the target conditions (e.g., IDS/consistent/social) as 1, and the alternative conditions (e.g., ADS/inconsistent/nonsocial) as 2.

```{r}
unique(df$tt_dir) # NA = 0, IDS = 1, ADS = 2
unique(df$tt_con) # NA = 0, consistent = 1, inconsistent = 2
unique(df$tt_soc) # NA = 0, social = 1, nonsocial = 2
```

<br>

#### Additional Exp:

```{r}
# As a "by" factor:
df$method <- as.factor(df$unique_subject_identifier)
df$participant_sex <- as.factor(df$participant_sex)
```

```{r}
# Exclusion: only if we would like to look into detailed stimulus types:
df$stimulus_id <- as.factor(df$stimulus_id)
```

```{r}
# Developmental factors:
df$age_days <- as.numeric(df$age_days)
df$age_group_new <- as.factor(df$age_group_new)
```


#### Random Var: `unique_subject_identifier` (factor)

*Exclusions*: The higher-level variable `lab_id` was no longer required.

```{r}
df$unique_subject_identifier <- as.factor(df$unique_subject_identifier)
```

<br>

### GAMM

```{r,include=FALSE}
library(mgcv)
library(itsadug)
```


```{r}
# Subset data for GAMM analyses
sub_df <- df[, c(
  # Y
  "looking_time",
  "looking_time2",
  # X
  "trial_number",
  # Exp
  "tt_dir", # MB1
  "tt_con", # MB3
  "tt_soc", # MB4
  # By
  "method",
  "participant_sex",
  # Dev
  "age_days",
  "age_group_new",
  #random
  "unique_subject_identifier"
  )]

# Sort data per individual trajectory
sub_df = sub_df[order(sub_df$unique_subject_identifier,sub_df$trial_number),]

# Mark the start of each trajectory (for ACF purposes)
sub_df$start.event <- sub_df$trial_number == 1

# Rename variables
names(sub_df) = c("y1","y2","x","dir","con","soc","method","gender","days","agegrp","id","start.event")

```

Still need hime hpc to run *echt GAMM*, but for the time being, use simple random effects `s(id, bs = "re")`. 

Autocorrelation not yet corrected!

#### Overall habituation trend

```{r}
setwd("/Users/yangji/Downloads/osfstorage-archive")
# m0 <- bam(y1 ~ s(x) + s(id, bs = "re"),
#              data = sub_df,
#              method = "fREML",
#              discrete = TRUE)
# saveRDS(m0, file="m0.rds")

m0 = readRDS("m0.rds")

rho0 = acf_resid(m0)[2] # no correction 

par(mfrow=c(1,2), cex=1.1)

plot_smooth(m0, view="x",
            xlim = c(1,12),
            xlab="Trial",ylab="LT (s)",
            rm.ranef=FALSE, main="Intercept + s(Trial) + RE", rug=TRUE,
            col='gray')

plot_smooth(m0, view="x",
            xlim = c(1,12),
            xlab="Trial",ylab="LT (s)",
            rm.ranef=TRUE, main="Intercept + s(Trial)", rug=TRUE,
            col='gray')
```

```{r}
# contrasts(sub_df$dir) <- 'contr.treatment'
# m1.1 <- bam(y1 ~ s(x) + dir + s(x, by = dir) + s(id, bs = "re"),
#             data = sub_df,
#             method = "fREML",
#             discrete = TRUE)
# saveRDS(m1.1, file="m1.1.rds")


m1.1 = readRDS("m1.1.rds")

rho0 = acf_resid(m1.1)[2] # no correction 

summary(m1.1)

# par(mfrow=c(2,2), cex=1.1)

plot_smooth(m1.1, view="x",cond=list(dir=0), 
            xlim = c(1,12),ylim = c(0,15),
            xlab="Trial",ylab="LT (s)",
            rm.ranef=TRUE, main="Intercept + s(Trial)", 
            col='gray') ## 0 = NA

plot_smooth(m1.1, view="x", cond=list(dir=1), 
            add=TRUE, col='green') ## 1 = IDS
plot_smooth(m1.1, view="x", cond=list(dir=2), 
            add=TRUE, col='red') ## 2 = ADS

plot_diff(m1.1, view="x", 
          comp=list(dir=c(1, 0)),
          # ylim = c(0,8),
          main="IDS - Baseline",
          col=c('blue','gray'), 
          ylab="Est. difference in LT (s)",
          rm.ranef=TRUE,sim.ci=TRUE)

plot_diff(m1.1, view="x", 
          comp=list(dir=c(2, 0)),
          # ylim = c(0,8),
          main="ADS - Baseline",
          col=c('blue','gray'),
          ylab="Est. difference in LT (s)",
          rm.ranef=TRUE,sim.ci=TRUE)

plot_diff(m1.1, view="x", 
          comp=list(dir=c(1, 2)),
          # ylim = c(0,8),
          main="IDS - ADS",
          col=c('blue','red'), 
          ylab="Est. difference in LT (s)",
          rm.ranef=TRUE,sim.ci=TRUE)

```



```{r}
# contrasts(sub_df$con) <- 'contr.treatment'
# m1.2 <- bam(y1 ~ s(x) + con + s(x, by = con) + s(id, bs = "re"),
#             data = sub_df,
#             method = "fREML",
#             discrete = TRUE)
# saveRDS(m1.2, file="m1.2.rds")

m1.2 = readRDS("m1.2.rds")

rho0 = acf_resid(m1.2)[2] # no correction 

summary(m1.2)

# par(mfrow=c(2,2), cex=1.1)

plot_smooth(m1.2, view="x",cond=list(con=0), 
            xlim = c(1,12),ylim = c(0,15),
            xlab="Trial",ylab="LT (s)",
            rm.ranef=TRUE, main="Intercept + s(Trial)", 
            col='gray') ## 0 = NA

plot_smooth(m1.2, view="x", cond=list(con=1), 
            add=TRUE, col='green') ## 1 = con
plot_smooth(m1.2, view="x", cond=list(con=2), 
            add=TRUE, col='red') ## 2 = incon

plot_diff(m1.2, view="x", 
          comp=list(con=c(1, 0)),
          # ylim = c(0,8),
          main="Consistent - Baseline",
          col=c('blue','gray'), 
          ylab="Est. difference in LT (s)",
          rm.ranef=TRUE,sim.ci=TRUE)

plot_diff(m1.2, view="x", 
          comp=list(con=c(2, 0)),
          # ylim = c(0,8),
          main="Inconsistent - Baseline",
          col=c('blue','gray'),
          ylab="Est. difference in LT (s)",
          rm.ranef=TRUE,sim.ci=TRUE)

plot_diff(m1.2, view="x", 
          comp=list(con=c(1, 2)),
          # ylim = c(0,8),
          main="Consistent - Inconsistent",
          col=c('blue','red'), 
          ylab="Est. difference in LT (s)",
          rm.ranef=TRUE,sim.ci=TRUE)
```

```{r}
# contrasts(sub_df$soc) <- 'contr.treatment'
# m1.3 <- bam(y1 ~ s(x) + soc + s(x, by = soc) + s(id, bs = "re"),
#             data = sub_df,
#             method = "fREML",
#             discrete = TRUE)
# saveRDS(m1.3, file="m1.3.rds")

m1.3 = readRDS("m1.3.rds")

rho0 = acf_resid(m1.3)[2] # no correction 

summary(m1.3)

# par(mfrow=c(2,2), cex=1.1)

plot_smooth(m1.3, view="x",cond=list(soc=0), 
            xlim = c(1,12),ylim = c(0,15),
            xlab="Trial",ylab="LT (s)",
            rm.ranef=TRUE, main="Intercept + s(Trial)",
            col='gray') ## 0 = NA

plot_smooth(m1.3, view="x", cond=list(soc=1), 
            add=TRUE, col='green') ## 1 = social
plot_smooth(m1.3, view="x", cond=list(soc=2), 
            add=TRUE, col='red') ## 2 = nonsocial

plot_diff(m1.3, view="x", 
          comp=list(soc=c(1, 0)),
          # ylim = c(0,8),
          main="Social - Baseline",
          col=c('blue','gray'), 
          ylab="Est. difference in LT (s)",
          rm.ranef=TRUE,sim.ci=TRUE)

plot_diff(m1.3, view="x", 
          comp=list(soc=c(2, 0)),
          # ylim = c(0,8),
          main="Nonsocial - Baseline",
          col=c('blue','gray'),
          ylab="Est. difference in LT (s)",
          rm.ranef=TRUE,sim.ci=TRUE)

plot_diff(m1.3, view="x", 
          comp=list(soc=c(1, 2)),
          # ylim = c(0,8),
          main="Social - Non-social",
          col=c('blue','red'), 
          ylab="Est. difference in LT (s)",
          rm.ranef=TRUE,sim.ci=TRUE)

```

#### Interactions (tensor est `te` only performed using *HPC*)

##### By method interactions

###### Differential trends across distinct methods

```{r}

# contrasts(sub_df$method) <- 'contr.treatment'
# 
# m2.1 <- bam(y1 ~ te(x, days, by = method) + method + s(id, bs = "re"),
#             data = sub_df,
#             method = "fREML",
#             discrete = TRUE,
#             trace = TRUE)
# saveRDS(m2.1, file="m2.1.rds")

# m2.1=readRDS("m2.1.rds")
# 
# summary(m2.1)
# plot_diff2(m2.1, view = c("x","method"))

```

###### Trends of various methods across directedness levels

```{r}

# sub_df$dir.m = interaction(sub_df$dir,sub_df$method)
# sub_df$dir.m = as.factor(sub_df$dir.m)
# 
# contrasts(sub_df$dir.m) <- 'contr.treatment'
# 
# m2.2 <- bam(y1 ~ te(x, days, by = dir.m) + dir.m + s(id, bs = "re"),
#             data = sub_df,
#             method = "fREML",
#             discrete = TRUE)
# saveRDS(m2.2, file="m2.2.rds")

# m2.2=readRDS("m2.2.rds")

```

###### Trends of various methods across consistency.

```{r}

# sub_df$con.m = interaction(sub_df$con,sub_df$method)
# sub_df$con.m = as.factor(sub_df$con.m)
# 
# contrasts(sub_df$con.m) <- 'contr.treatment'
# 
# m2.3 <- bam(y1 ~ te(x, days, by = con.m) + con.m + s(id, bs = "re"),
#             data = sub_df,
#             method = "fREML",
#             discrete = TRUE)
# saveRDS(m2.3, file="m2.3.rds")

# m2.3=readRDS("m2.3.rds")

```

###### Trends of various methods across levels of social interaction.

```{r}

# sub_df$soc.m = interaction(sub_df$soc,sub_df$method)
# sub_df$soc.m = as.factor(sub_df$soc.m)
# 
# contrasts(sub_df$soc.m) <- 'contr.treatment'
# 
# m2.4 <- bam(y1 ~ te(x, days, by = soc.m) + soc.m + s(id, bs = "re"),
#             data = sub_df,
#             method = "fREML",
#             discrete = TRUE)
# saveRDS(m2.4, file="m2.4.rds")

# m2.4=readRDS("m2.4.rds")

```


##### By gender interactions

###### Differential trends across genders

```{r}

# contrasts(sub_df$method) <- 'contr.treatment'
# 
# m3.1 <- bam(y1 ~ te(x, days, by = gender) + gender + s(id, bs = "re"),
#             data = sub_df,
#             method = "fREML",
#             discrete = TRUE,
#             nthreads = 8)
# saveRDS(m3.1, file="m3.1.rds")

# m3.1=readRDS("m3.1.rds")

```

###### Trends of genders across directedness levels

```{r}

# sub_df$dir.m = interaction(sub_df$dir,sub_df$gender)
# sub_df$dir.m = as.factor(sub_df$dir.m)
# 
# contrasts(sub_df$dir.m) <- 'contr.treatment'
# 
# m3.2 <- bam(y1 ~ te(x, days, by = dir.m) + dir.m + s(id, bs = "re"),
#             data = sub_df,
#             method = "fREML",
#             discrete = TRUE)
# saveRDS(m3.2, file="m3.2.rds")

# m3.2=readRDS("m3.2.rds")

```

###### Trends of genders across consistency.

```{r}

# sub_df$con.m = interaction(sub_df$con,sub_df$gender)
# sub_df$con.m = as.factor(sub_df$con.m)
# 
# contrasts(sub_df$con.m) <- 'contr.treatment'
# 
# m3.3 <- bam(y1 ~ te(x, days, by = con.m) + con.m + s(id, bs = "re"),
#             data = sub_df,
#             method = "fREML",
#             discrete = TRUE)
# saveRDS(m3.3, file="m3.3.rds")

# m3.3=readRDS("m3.3.rds")

```

###### Trends of genders across levels of social interaction.

```{r}

# sub_df$soc.m = interaction(sub_df$soc,sub_df$gender)
# sub_df$soc.m = as.factor(sub_df$soc.m)
# 
# contrasts(sub_df$soc.m) <- 'contr.treatment'
# 
# m3.4 <- bam(y1 ~ te(x, days, by = soc.m) + soc.m + s(id, bs = "re"),
#             data = sub_df,
#             method = "fREML",
#             discrete = TRUE)
# saveRDS(m3.4, file="m3.4.rds")

# m3.4=readRDS("m3.4.rds")

```


